---
title: "Introduction to openalexR: a webinar"
date: today
author: Trang Le
format: 
  html:
    theme: 
      light: sandstone
      dark: slate
    toc: true
    toc_float: true
    resources:
      - data/* 
# filters:
#   - webr
---

Welcome to the webinar on **openalexR**!
Today, we will explore how to use the **openalexR** package to fetch data from OpenAlex, a free, comprehensive database for scholarly papers, authors, institutions, and more.

## 🌻 Introduction

**openalexR** is an R package that provides an interface to OpenAlex, a free and open dataset of the entire research landscape.
With **openalexR**, you can easily access and analyze scholarly data directly from R.

## 🌱 Installation and setup

If you haven't already installed the **openalexR** package, you can do so from CRAN:

```{r install, eval=FALSE}
install.packages("openalexR")
```

::: callout-important
## The polite pool

Before we go any further, we highly recommend you set the `openalexR.mailto` option so that your requests go to [the polite pool](https://docs.openalex.org/how-to-use-the-api/rate-limits-and-authentication#the-polite-pool) for faster response times.
If you have OpenAlex Premium, you can add your API key to the `openalexR.apikey` option as well.
To do so, you can open `.Renviron` with `file.edit("~/.Renviron")` and add:

```         
openalexR.mailto = example@email.com
openalexR.apikey = EXAMPLE_APIKEY
```
:::

We will now load in **openalexR** and two additional packages **dplyr**, and **ggplot2** to use for the rest of this webinar.

```{r load, message=FALSE}
library(openalexR)
library(dplyr)
library(ggplot2)
```

```{r set-theme, include=FALSE}
theme_set(theme_classic())
theme_update(
  plot.background = element_rect(fill = "transparent", colour = NA),
  panel.background = element_rect(fill = "transparent", colour = NA)
)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))
```

::: {.callout-note collapse="true"}
## R base pipe

Throughout the webinar, you will see this symbol, `|>`.
Simply put, `x |> f()` is equivalent to `f(x)`.
Using this [R base pipe](https://developer.r-project.org/Blog/public/2021/02/21/a-native-pipe-in-r-2021-02-21/) allows me to chain functions together in a more readable way.
:::

## 🌿 Basic usage

The main function of **openalexR** is `oa_fetch`.

```{r}
#| eval: false
?oa_fetch
```

### Fetching information from identifiers

Let's start by fetching data on a specific scholarly work using its OpenAlex ID:

```{r fetch-work}
work <- oa_fetch(entity = "works", id = "W2741809807", verbose = TRUE)
work
```

::: {.callout-note collapse="true"}
When you know the OpenAlex ID of the work, you do not need to specify `entity = "works"` because entity can be inferred from the first character of `id`.
However, I specify it here for clarity.
In other use cases (search, filter), you will almost always want to specify `entity`.
:::

Now, we can view the output tibble/dataframe, `work`, interactively in RStudio or inspect it with base functions like `str` or `head`.

```{r}
str(work, max.level = 2)
```

**openalexR** also provides the `show_works` function to simplify the result (e.g., remove some columns, keep first/last author) for easy viewing.
Let us define `print_oa` to wrap the output table in `knitr::kable()` to be displayed nicely on the webpage, but you will most likely not need this function.

```{r}
print_oa <- function(x, fun = show_works) {
  x |>
    select(-any_of("url")) |>
    fun() |>
    knitr::kable()
}
print_oa(work)
```

### Filter and search

There are different [filters](https://ropensci.github.io/openalexR/articles/Filters)/arguments you can use in `oa_fetch`, depending on which [entity](https://docs.openalex.org/#data) you're interested in: works, authors, sources, funders, institutions, or concepts.
We show a few examples below.

#### 1. Use `doi` as a [works filter](https://ropensci.github.io/openalexR/articles/Filters.html#works)

**Goal**: Fetch information on two works with DOIs `10.1016/j.joi.2017.08.007` and `https://doi.org/10.1007/s11192-013-1221-3`.
[^1]

[^1]: Either canonical form with [https://doi.org/](https://orcid.org/){.uri} or without will work.

```{r}
works_from_dois <- oa_fetch(
  entity = "works",
  doi = c("10.1016/j.joi.2017.08.007", "https://doi.org/10.1007/s11192-013-1221-3"),
  verbose = TRUE
)
print_oa(works_from_dois)
```

#### 2. Use `author.orcid` as a filter

**Goal**: Fetch all works by authors with ORCID IDs `0000-0001-6187-6610` and `0000-0002-8517-9411`.
[^2]

[^2]: Either canonical form with <https://orcid.org/> or without will work.

```{r}
works_from_orcids <- oa_fetch(
  entity = "works",
  author.orcid = c("0000-0001-6187-6610", "0000-0002-8517-9411")
)

print_oa(works_from_orcids)
```

#### 3. Many filters at once

**Goal**: Download all works that have been cited more than 50 times, published between 2020 and 2021, and include the strings "bibliometric analysis" or "science mapping" in the title.
Maybe we also want the results to be sorted by total citations in a descending order.

```{r}
works_biblio <- oa_fetch(
  entity = "works",
  title.search = c("bibliometric analysis", "science mapping"),
  cited_by_count = ">50",
  from_publication_date = "2020-07-01",
  to_publication_date = "2021-07-01",
  options = list(sort = "cited_by_count:desc")
)

print_oa(works_biblio)
```

#### 4. What if we use a wrong filter?

**Goal**: Sample 10 works published in 2024.

Say we mistakenly use `year` instead of `publication_year` as a filter.

```{r}
#| error: true
oa_fetch(
  entity = "works",
  year = 2024,
  options = list(sample = 10),
  verbose = TRUE
)
```

The API returns a helpful error message: `year` is a wrong filter, and shows you all possible, correct filters.
In our case, `publication_year` is the filter we want.

```{r}
oa_fetch(
  entity = "works",
  publication_year = 2024,
  # TODO figure out why this doesn't work with some seeds
  options = list(sample = 10, seed = 1),
  verbose = TRUE
) |>
  print_oa()
```

#### 5. Search

You can also search for works related to a specific topic.
Here's an example fetching the first few works of which abstracts include the phrase "urban heat island":

```{r search-works}
oa_fetch(
  entity = "works",
  abstract.search = "urban heat island",
  options = list(sample = 10, seed = 1)
) |>
  print_oa()
```

::: {.callout-tip collapse="true"}
You can read more about search [here](https://docs.openalex.org/how-to-use-the-api/get-lists-of-entities/search-entities).
<!-- You will learn how relevance score is calculated, how words are stemmed to improve search results, and how to do complex boolean searches. --> Specifically, stop words like "the" and "an" are removed, and, by default, [stemming](https://www.elastic.co/guide/en/elasticsearch/reference/current/analysis-kstem-tokenfilter.html) is used to improve results ("possums" will also return records with the word "possum").
To disable stemming and the removal of stop words for searches on titles and abstracts, you can add `.no_stem` to the search filter, *e.g.* `abstract.search.no_stem`.
:::

#### 6. Other entities

Similar to works, you can fetch information on authors, sources, funders, institutions, and concepts.
Here are some examples of fetching authors:

**Goal**: Acquire information on a couple of authors with more than 10 works.
Here, we can use filters such as `display_name` and `works_count`:

```{r}
oa_fetch(
  entity = "authors",
  display_name.search = c("Massimo Aria", "Eugen Lounkine"),
  works_count = ">10"
) |>
  print_oa(show_authors)
```

## 🪴 Example analyses

**Goal**: Download all authors' records of scholars who work at the [University of Naples Federico II](https://explore.openalex.org/institutions/I71267560) (OpenAlex ID: I71267560) and have published at least 500 publications.

Let's first check how many records match the query, then download the entire collection.
We can do this by adding `count_only` (default `FALSE`) to the list of argument:

```{r}
oa_fetch(
  entity = "authors",
  last_known_institutions.id = "I71267560",
  works_count = ">499",
  count_only = TRUE
)

oa_fetch(
  entity = "authors",
  last_known_institutions.id = "I71267560",
  works_count = ">499"
) |>
  print_oa(show_authors) 
```

TODO

## 🌵 Advanced usage

### Other parameters of `oa_fetch`

So far, we have seen the argument `options`, which is a list of additional parameters that can be passed to the API.
`output = "list"` TODO

### Building your own query

TODO

### oa_generate

TODO

### Analyzing Citations

Analyze the citations of a work:

```{r analyze-citations}
# citations <- oa_fetch(entity = "works", filter = list(cited_by.work.id = work_id))
# print(citations)
```

TODO

## 🛤️ Coding challenges

### Challenge 1: Advanced filters

According to the OpenAlex [API documentation](https://docs.openalex.org/api-entities), what is the filter we should use to:

-   Get `funders` with a description containing "engineering"?
-   Get `topics` with more than 1000 works?
-   Get `institutions` that are located in Asia?

### Challenge 2: Humpback whale

Identify works on a specific topic (e.g., "humpback whale") that have been cited more than 100 times.
When were these works published?
Where are the authors based?
Create a bar plot showing the number of citations for each author.

::: panel-tabset
## Your solution

```{r challenge-top-authors}
#| eval: false

work <- oa_fetch(
  entity = "works",
  ______
)

# print(humpback$title[1:10])
# hb_authors <- humpback$author |> 
#   bind_rows() 

```

## My solution

```{r answer-top-authors}
humpback <- oa_fetch(
  entity = "works",
  title.search = "humpback whale",
  cited_by_count = ">100",
  options = list(sort = "cited_by_count:desc")
)
print(humpback$title[1:10])
hb_authors <- humpback$author |> 
  bind_rows() 
```
:::

## 🌸 Q&A

TODO

# 🍁 Conclusion

Thank you for participating in the **openalexR** webinar!
We hope you found the session informative and engaging.
For more information, visit the [openalexR GitHub page](https://github.com/ropensci/openalexR) reach out to me on the [OpenAlex Community](https://groups.google.com/g/openalex-community) Google Group with any questions.

------------------------------------------------------------------------

## 🌳 Links

-   <https://openalex.org>
-   <https://docs.openalex.org>
-   <https://docs.ropensci.org/openalexR>
